Liveness-Based Garbage Collection for Lazy Languages

### 2016-02-06 paper/paper.tex
Changed as per discussion on 5th Feb

\todelete{While  this  idea has  been  shown  to  be effective  for  a
  first-order     {\em     eager}    language~\cite{asati14lgc},     a
  straightforward extension of the technique  is not possible for lazy
  languages,  where   heap-allocated  objects  may   include  closures
  (runtime   representations    of   unevaluated    expressions).    }
\added{Designing a liveness based garbage collector for lazy languages
  poses significant challenges.}  Firstly, since  data is made live by
evaluation of closures, and in lazy languages the place in the program
where  this evaluation  takes place  cannot be  statically determined,
laziness   complicates  liveness   analysis  itself.    Moreover,  for
liveness-based GC  to be effective, we  need to extend
it to closures apart from evaluated  data.  Since a closure can escape
the  scope in  which it  was  created, it  has to  carry the  liveness
information  of  its  free  variables.  As  execution  progresses  and
possible  future   uses  are   eliminated,  we  update   the  liveness
information  in a  closure with  a more  precise version.   Because of
these reasons,  the garbage collector also  becomes significantly more
complicated than a liveness-based collector for a eager language.

@@

In  this work  we show  that  static analysis  of heap  data can  help
garbage  collectors   in  reclaiming   more  garbage.    The  specific
contributions of this paper are
\added{
\begin{itemize}
\item We propose a liveness based GC scheme for a lazy
  first order functional  language and prove its  correctness.  To the
  best  of our  knowledge, this  is the  first work  on liveness-based
  GC   for    lazy    languages.    All    previous
  attempts~\cite{shaham01heap,  ran.shaham-sas03,  shaham02estimating,
    asati14lgc,    karkare06effectiveness}   have    involved   either
  imperative  or  eager  functional  languages.   The  novelty  of  our
  approach is in  handling of  liveness of  functional values  (i.e.,
  closures), an issue that arises due to laziness.

@@

\item The  analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  whether to  copy a  cell during GC translates  to a
  membership problem for such grammars.   In this paper, we have shown
  the undecidability  of this  problem. We  circumvent the  problem by
  over-approximating the grammars with DFAs.

@@

%\section{Computing liveness and its encoding as DFA}\label{sec:computing}
\section{Generating and storing liveness information}\label{sec:computing}
Section~\ref{sec:liveness}  gave   a  fully context-sensitive
liveness analysis which 
%$\mathcal{L}$ and $ref$ together
described the  liveness set for  each variable  in a function  body in
terms of  a symbolic demand  $\sigma$ and \Lfonly.  However,  we are yet
to describe  (i) how to obtain demand  transformers \Lfonly\ from
the  rule {\sc  live-define}, and  (ii)  how to  compute the  concrete
demand $\sigma$  on each function.   To do so,  we first need to  modify the
liveness rules to a different form.

%\subsection{Modifying liveness rules}
\subsection{Tweaking the rules}

@@

%\subsection{Generating equations for $\Lfonly_{\mathit f}$}
\subsection{Computing function summaries $\Lfonly_{\mathit f}$}
Given a  function $\mathit{f}$, we now  describe how to
generate  equations   for  the   demand  transformation
\Lfonly$_\mathit{f}$.        The       program       in
Figure~\ref{fig:mot-example2-a}   serves  as   a  running
example.  Starting with a  symbolic demand $\sigma$, we
determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In
particular, we  consider $\Lanv{x_i}{}$,  the liveness
of the $i^{\mbox{\footnotesize th}}$ parameter $x_i$.   By   the   rule   {\sc
  live-define},   this   should    be   the   same   as
$\Lf{f}{i}{\sigma}$. Applying this to \length, we have:
$$
 \Lf{\length}{1}{\sigma} = \Lanv{\xl}{} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
  \cup \clazy\Lf{\length}{1}{\clazy\sigma}
$$

@@

%\subsection{Generating liveness equations \Lv\  for function bodies}
\subsection{Handling user-defined functions}
\label{sec:bodylivenessbodies}

@@

In the  space of  functional languages,  there are:  rewriting methods
such                                                                as
deforestation~\cite{wadler88deforest,gill93ashort,chitil99deforest},
sharing  analysis   based  reallocation~\cite{jones89compile},  region
based  analysis~\cite{tofte98region},  and insertion  of  compile-time
nullifying     statements~\cite{inoue88analysis,lee05static,Hamilton}.
All compile-time marking  approaches rely on an  efficient and precise
alias  analysis  and cannot  provide  significant  improvement in  its
absence.  An interesting approach suggested in~\cite{HofmannJ03} is to
annotate the heap usage of  first-order programs through linear types.
The  annotations  are  then  used to  serve  memory  requests  through
re-allocation.  However, this requires the user to write programs in a
specific way. 
%
Safe-for-space~\cite{appel.cps}              approaches~\cite{Clinger,
  shao00efficient}  reduce the  amount of  heap used  by a  program by
allocating   closures    in   registers   and   through    tail   call
optimizations. However, these approaches take care of only part of the
problem addressed  by our  analysis as the  program can  still contain
unused objects and closures that are reachable.
%

>>>>

\added{While  the idea  of  using static  analysis  to improve  memory
  utilization  has  been  shown  to   be  effective  for  {\em  eager}
  languages~\cite{asati14lgc,       HofmannJ03,       inoue88analysis,
    lee05static}, a straightforward extension  of the technique is not
  possible  for  lazy  languages,  where  heap-allocated  objects  may
  include   closures    (runtime   representations    of   unevaluated
  expressions).   }   \todelete{Designing  a  liveness  based  garbage
  collector   for  lazy   languages  poses   significant  challenges.}
Firstly, since  data is made  live by  evaluation of closures,  and in
lazy languages  the place in  the program where this  evaluation takes
place cannot  be statically determined, laziness  complicates liveness
analysis itself.  Moreover, for liveness-based  GC to be effective, we
need to  extend it  to closures  apart from  evaluated data.   Since a
closure can escape the scope in which  it was created, it has to carry
the  liveness  information  of   its  free  variables.   As  execution
progresses  and possible  future uses  are eliminated,  we update  the
liveness  information  in  a  closure with  a  more  precise  version.
Because  of   these  reasons,  the  garbage   collector  also  becomes
significantly more  complicated than a liveness-based  collector for a
eager language.

@@

In  this work  we show  that  static analysis  of heap  data can  help
garbage  collectors   in  reclaiming   more  garbage.    The  specific
contributions of this paper are
\added{
\begin{itemize}
\item We  propose a liveness  based GC scheme  for a lazy  first order
  functional language and  prove its correctness.  To the  best of our
  knowledge,  this is  the first  work on  liveness-based GC  for lazy
  languages.       All      previous      attempts~\cite{shaham01heap,
    ran.shaham-sas03,          shaham02estimating,         asati14lgc,
    karkare06effectiveness} have  involved either imperative  or eager
  functional languages.  The novelty of our approach is in handling of
  liveness  of closures,  an  issue  that arises  due  to laziness  at
  runtime.

@@

\item The  analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  whether to copy a cell during  GC translates to a membership problem
  for  such  grammars.  Earlier  assumed  the  undecidability of  this
  membership question  and used an over-approximation  to overcome it.
  In this  paper, we provide a  formal proof of the  undecidability of
  this problem.

@@

\section{Computing liveness and its encoding as DFA}\label{sec:computing}
%\section{Generating and storing liveness information}\label{sec:computing}
Section~\ref{sec:liveness}  gave   a  fully context-sensitive
liveness analysis which 
%$\mathcal{L}$ and $ref$ together
described the  liveness set for  each variable  in a function  body in
terms of  a symbolic demand  $\sigma$ and \Lfonly.  However,  we are yet
to describe  (i) how to obtain demand  transformers \Lfonly\ from
the  rule {\sc  live-define}, and  (ii)  how to  compute the  concrete
demand $\sigma$  on each function.   To do so,  we first need to  modify the
liveness rules to a different form.

%\subsection{Modifying liveness rules}
\paragraph{Symbolic representation of operations}

@@

%\subsection{Generating equations for $\Lfonly_{\mathit f}$}
\paragraph{Computing function summaries $\Lfonly_{\mathit f}$}
Given a  function $\mathit{f}$, we now  describe how to
generate  equations   for  the   demand  transformation
\Lfonly$_\mathit{f}$.        The       program       in
Figure~\ref{fig:mot-example2-a}   serves  as   a  running
example.  Starting with a  symbolic demand $\sigma$, we
determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In
particular, we  consider $\Lanv{x_i}{}$,  the liveness
of the $i^{\mbox{\footnotesize th}}$ parameter $x_i$.   By   the   rule   {\sc
  live-define},   this   should    be   the   same   as
$\Lf{f}{i}{\sigma}$. Applying this to \length, we have:
$$
 \Lf{\length}{1}{\sigma} = \Lanv{\xl}{} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
  \cup \clazy\Lf{\length}{1}{\clazy\sigma}
$$

@@

%\subsection{Generating liveness equations \Lv\  for function bodies}
\paragraph{Handling user-defined functions}
\label{sec:bodylivenessbodies}

@@

In the space of functional languages, there are: rewriting methods
such as
deforestation~\cite{wadler88deforest,gill93ashort,chitil99deforest},
sharing analysis based reallocation~\cite{jones89compile}, region
based analysis~\cite{tofte98region}, and insertion of compile-time
nullifying statements~\cite{inoue88analysis}.  All compile-time
marking approaches rely on an efficient and precise alias analysis and
cannot provide significant improvement in its absence. The only work
in the space of lazy languages seems to be~\cite{lee05static,Hamilton}
which touches upon only basic techniques of compile-time garbage
marking, explicit deallocation and desctructive allocation. An
interesting approach suggested in~\cite{HofmannJ03} is to annotate the
heap usage of first-order programs through linear types.  The
annotations are then used to serve memory requests through
re-allocation.  However, this requires the user to write programs in a
specific way.
%
Safe-for-space~\cite{appel.cps}              approaches~\cite{Clinger,
  shao00efficient}  reduce the  amount of  heap used  by a  program by
allocating   closures    in   registers   and   through    tail   call
optimizations. However, these approaches take care of only part of the
problem addressed  by our  analysis as the  program can  still contain
unused objects and closures that are reachable.
%
